---
title: "MSstatsResponse: A package for detecting drug-protein interactions in dose-response mass spectrometry-based proteomics experiments"
author: "Sarah Szvetecz (<szvetecz.s@northeastern.edu>)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MSstatsResponse User Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

MSstatsResponse provides statistical tools for detecting drug-protein interactions and estimating IC50 values from dose-response mass spectrometry-based proteomics experiments. The package implements isotonic regression models to detect monotonic dose-response relationships and provides robust statistical inference through F-tests and bootstrap confidence intervals.

MSstatsResponse is designed to work seamlessly with summarized data from MSstats and MSstatsTMT workflows, but can also process other dose-response proteomics datasets that include protein-level quantifications across drug treatments.

### Statistical Framework

The package employs isotonic regression, a non-parametric method that fits monotonic functions without assuming a specific functional form. This approach is particularly suitable for dose-response data where monotonicity (increasing or decreasing trend) is expected.

### Analysis Workflow

MSstatsResponse implements a three-step statistical analysis pipeline:

0. **Data Preparation**: `MSstatsPrepareDoseResponseFit()` - Convert input data to standardized format
1. **Interaction Detection**: `DoseResponseFit()` - Detect drug-protein interactions using isotonic regression and F-tests  
2. **IC50 Estimation**: `PredictIC50()` - Estimate IC50 values with bootstrap confidence intervals
3. **Power Analysis for Experimental Planning**: `FutureExperimentSimulation` - Simulate dose-response experiments to determine optimal study designs and predict true-positive and false-positive rates

 

## Installation and Setup

### Install MSstatsResponse

```{r install, eval=FALSE}
# Install from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("MSstatsResponse")
```

### Load Required Libraries

```{r load_libraries, message=FALSE}
library(MSstatsResponse)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(data.table)

# Optional: for upstream data processing
# library(MSstats)
# library(MSstatsTMT)
```

 

## 1. Data Preprocessing with MSstats

Before using MSstatsResponse, data should be processed through standard MSstats workflows for cleaning, normalization, and protein summarization.

### Example MSstats Workflow

```{r msstats_preprocessing, eval=FALSE}
# Read raw data (example with Spectronaut output)
raw_data <- read_tsv("path/to/spectronaut_report.tsv")

# Convert to MSstats format
msstats_data <- SpectronauttoMSstatsFormat(raw_data)

# Process data: normalization and protein summarization
processed_data <- dataProcess(
  msstats_data,
  normalization = "equalizeMedians",  # or FALSE for no normalization
  summaryMethod = "TMP",              # Tukey's median polish
  MBimpute = TRUE,                     # Impute missing values
  maxQuantileforCensored = 0.999
)

# Extract protein-level data for dose-response analysis
protein_level_data <- processed_data$ProteinLevelData
```

### Loading Example Data

For this vignette, we'll use pre-processed example data:

```{r load_example_data}
# Load pre-processed DIA-MS data example
data_path <- system.file("extdata", "DIA_MSstats_Normalized.RDS", 
                         package = "MSstatsResponse")
dia_normalized <- readRDS(data_path)

# Examine data structure
str(dia_normalized$ProteinLevelData[1:5, ])
```

 

## 2. Data Preparation for Dose-Response Analysis

### Converting GROUP Labels to Numeric Doses

The `ConvertGroupToNumericDose()` function parses MSstats-style GROUP labels to extract drug names and numeric dose values.

**Format expectations:**  
- Control samples: `"DMSO"` (assigned dose = 0)  
- Treatment samples: `"DrugName_ValueUnit"` (e.g., `"Dasatinib_003uM"`, `"Capmatinib_300nM"`) 

```{r convert_doses}
# Convert GROUP labels to numeric doses
dose_info <- ConvertGroupToNumericDose(dia_normalized$ProteinLevelData$GROUP)

# Add dose and drug information to the dataset
dia_normalized$ProteinLevelData <- dia_normalized$ProteinLevelData %>%
  mutate(
    dose_nM = dose_info$dose_nM,  # Dose in nanomolar
    drug = dose_info$drug          # Drug name
  )


# View converted data
dia_normalized$ProteinLevelData %>%
  select(Protein, GROUP, drug, dose_nM) %>%
  head(10)
```

### Formatting Data with MSstatsPrepareDoseResponseFit()

This function standardizes the data structure for downstream analysis:

```{r prepare_data}
# Prepare data for dose-response fitting
dia_prepared <- MSstatsPrepareDoseResponseFit(
  data = dia_normalized$ProteinLevelData,
  dose_column = "dose_nM",
  drug_column = "drug",
  protein_column = "Protein",
  log_abundance_column = "LogIntensities",
  transform_nM_to_M = TRUE  
)

# Examine prepared data structure
str(dia_prepared)

# View sample of prepared data
dia_prepared %>%
  filter(protein %in% c("PROTEIN_A", "PROTEIN_B")) %>%
  arrange(protein, drug, dose) %>%
  head(20)
```

 

## 3. Drug-Protein Interaction Detection

### Understanding Isotonic Regression

Isotonic regression fits a monotonic function to the data without assuming a parametric form. For this example, we have drug inhibitor dose-response data. When treated with inhibitors we typically expect:  

- **Non-increasing response**: Drug binding reduces protein abundance  
- **Monotonic constraint**: Response doesn't increase with dose

### Running DoseResponseFit()

The `DoseResponseFit()` function:  
1. Fits isotonic regression for each protein-drug pair  
2. Performs F-test comparing the fitted model to a flat (null) model  
3. Adjusts p-values for multiple testing using Benjamini-Hochberg procedure  


```{r fit_dose_response, message=FALSE, warning=FALSE}
# Detect drug-protein interactions
response_results <- DoseResponseFit(
  data = dia_prepared,
  increasing = FALSE,        # Expect decreasing response
  transform_dose = TRUE,     # Apply log10(dose + 1) transformation
  ratio_response = FALSE     # Stay on log2 scale for testing
)

# Examine results
response_results %>%
  select(protein, drug, F_statistic, P_value, adjust_pval) %>%
  arrange(adjust_pval) %>%
  head(10)
```

### Interpreting Results

- **F_statistic**: Ratio of between-dose to within-dose variance  

- **P_value**: Raw p-value from F-test  

- **adjust_pval**: FDR-adjusted p-value (Benjamini-Hochberg)  


Proteins with `adjust_pval < 0.05` show significant dose-dependent responses.  


```{r summarize_results}
# Count significant interactions
n_total <- nrow(response_results)
n_significant <- sum(response_results$adjust_pval < 0.05)

cat("Total protein-drug pairs tested:", n_total, "\n")
cat("Significant interactions (FDR < 0.05):", n_significant, "\n")
cat("Percentage significant:", round(100 * n_significant/n_total, 1), "%\n")

# Top hits
top_hits <- response_results %>%
  filter(adjust_pval < 0.05) %>%
  arrange(adjust_pval) %>%
  head(5)

print(top_hits)
```

 

## 4. IC50 Estimation

### Understanding IC50 Calculation

IC50 represents the dose at which the response is reduced to 50% of the control (DMSO) level. MSstatsResponse estimates IC50 by:  

1. Fitting isotonic regression on the ratio scale (response relative to DMSO)  

2. Finding the dose where fitted response = 0.5  

3. Computing confidence intervals via bootstrap resampling  


### Scale Considerations

- **Log2 scale**: Best for interaction testing (stabilizes variance)
- **Ratio scale**: Best for IC50 estimation (interpretable as fraction of control)

### Running PredictIC50()

```{r predict_ic50, message=FALSE, warning=FALSE}
# Estimate IC50 with bootstrap confidence intervals
ic50_predictions <- PredictIC50(
  data = dia_prepared,
  ratio_response = TRUE,     # Use ratio scale for IC50
  transform_dose = TRUE,     # Log-transform doses
  increasing = FALSE,        # Decreasing response expected
  bootstrap = TRUE,          # Compute confidence intervals
  n_samples = 1000,         # Number of bootstrap samples
  alpha = 0.10              # 90% confidence intervals
)

# View IC50 estimates
ic50_predictions %>%
  arrange(IC50) %>%
  head(10)
```


### Parallel Processing for Large Datasets

For datasets with many proteins, use parallel processing:

```{r parallel_ic50, eval=FALSE}
# Parallel IC50 estimation (faster for large datasets)
ic50_parallel <- PredictIC50Parallel(
  data = dia_prepared,
  ratio_response = TRUE,
  transform_dose = TRUE,
  bootstrap = TRUE,
  n_samples = 1000,
  numberOfCores = 4  # Use 4 CPU cores
)
```

 

## 5. Visualization

### Individual Protein Dose-Response Curves

The `VisualizeResponseProtein()` function creates publication-quality dose-response plots:

```{r visualize_single, fig.height=5, fig.width=7}
# Visualize strong responder
VisualizeResponseProtein(
  data = dia_prepared,
  protein_name = "PROTEIN_A",
  drug_name = "Drug1",
  ratio_response = TRUE,
  show_ic50 = TRUE,
  add_ci = TRUE,
  transform_dose = TRUE,
  n_samples = 1000
)

VisualizeResponseProtein(
  data = dia_prepared,
  protein_name = "PROTEIN_A",
  drug_name = "Drug1",
  ratio_response = TRUE,
  show_ic50 = TRUE,
  add_ci = TRUE,
  transform_dose = TRUE,
  n_samples = 1000
)
```

```{r visualize_another, fig.height=5, fig.width=7}
# Visualize another target
VisualizeResponseProtein(
  data = dia_prepared,
  protein_name = "PROTEIN_B",
  drug_name = "Drug1",
  ratio_response = TRUE,
  show_ic50 = TRUE,
  add_ci = TRUE
)
```

### Comparing Log2 vs Ratio Scale Visualization

```{r compare_scales, fig.height=4, fig.width=12}
# Log2 scale (left panel)
p1 <- VisualizeResponseProtein(
  data = dia_prepared,
  protein_name = "PROTEIN_A",
  drug_name = "Drug1",
  ratio_response = FALSE,  # Log2 scale
  show_ic50 = TRUE,
  add_ci = FALSE
)

# Ratio scale (right panel)  
p2 <- VisualizeResponseProtein(
  data = dia_prepared,
  protein_name = "PROTEIN_A",
  drug_name = "Drug1",
  ratio_response = TRUE,   # Ratio scale
  show_ic50 = TRUE,
  add_ci = FALSE
)

# Combine plots (requires gridExtra)
 gridExtra::grid.arrange(p1, p2, ncol = 2)
```

 

## 6. Experimental Design Planning

### Simulating Future Experiments

The `FutureExperimentSimulation()` function helps optimize experimental design by simulating performance under different conditions:

### Using Your Own Data as Templates. 

You can use proteins from your own experimental data as templates for more realistic simulations. This approach leverages your validated hits to model expected performance:
```{r}
# First, identify proteins from your results to use as templates
# Run simulation using your data as templates

custom_simulation <- FutureExperimentSimulation(
  N_proteins = 3000,
  N_rep = 2,
  N_Control_Rep = 3,
  Concentrations = c(0, 1, 3, 10, 30, 100, 300, 1000, 3000),
  data = dia_prepared,                          # Your data
  strong_proteins = 'PROTEIN_B',                # User defined strong responder
  weak_proteins = 'PROTEIN_A',                    # User defined weak responder 
  no_interaction_proteins = 'PROTEIN_C',          # User defined negative control 
  drug_name = "Drug1",                     # Specify which drug to model
  IC50_Prediction = FALSE
)

# Compare performance metrics
print(custom_simulation$Hit_Rates_Plot)

# Examine the templates that were extracted from your data
print(custom_simulation$Template_Used)
```


## Session Information

```{r session_info}
sessionInfo()
```

 

## References

1. Szvetecz, S., et al. (2025).MSstatsResponse: Semi-parametric statistical model enhances drug-protein interaction detection and IC50 estimation in chemoproteomics experiments. *Manuscript in preparation.*  

2. Choi, M., et al. (2014). MSstats: an R package for statistical analysis of quantitative mass spectrometry-based proteomic experiments. *Bioinformatics*, 30(17), 2524-2526.

